Computing in the Field of Rationals

### 2012-02-27 rational.tex
lots of edits

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

@@

@@

{{eta}}

\begin{center}
  \includegraphics{diagrams/eta.pdf}
\end{center}

Note that the connective is a {{+}}, hence only one of the branches
may be inhabited at any time. Thus the action of {{eta}} is to
transfer a backward flowing value on one wire to a forward flowing
value on the other wire.

{{eps}}

\begin{center}
  \includegraphics{diagrams/eps.pdf}
\end{center}

Coherence condition

@@

@@

\section{Semantics}
%%%%%%%%%%%%%%%
\subsection{Syntax}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions in {{langRevEE}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions Specific to Negation}

>>>>

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^*
%! eps* = \epsilon^*
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{ {{langRev}} }

The presentation here differs from \cite{infeffects} in the sense that
we have introduced the {{0}} type here, thereby completeing the monoid
{{(0,+)}} and we present a small-step operational semantics.

%subcode{bnf} include main
% Combinators, c = iso | c (;) c | c (*) c | c (+) c
% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
%                  &|& LeftTimes cc c v | RightTimes c v cc 
%                  &|& LeftPlus cc c | RightPlus c cc 
% Values, v = () | (v, v) | L v | R v
%
% Machine states = <c, v, cc> | {[c, v, cc]}
% Start state = <c, v, []> 
% Stop State = {[c, v, []]}

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}

@@

Additive duality is added to {{langRev}} by addition of negative types
denoted {{-b}} and the isomorphisms {{eta+}} and {{eps+}}. Hence we
extend the language as follows:

%subcode{bnf} include main
% Base Types, b = ... | -b
% Values, v = ... | -v
%
% Isomorphisms, iso &=& ... | eta+ | eps+

\noindent
with the types judgements

%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+

We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
dualy on the right is {{eps+}} showing the map from {{-b+b}} to
0. Even though the diagrams below show a {{0}} wire for completeness,
later diagrams will always drop them in contexts where we can
implicitely introduce/eliminate then using {{zeroi}}/{{zeroe}}.

\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/eta.pdf}
\end{center}

\begin{center}
  \includegraphics{diagrams/eps.pdf}
\end{center}
\end{multicols}

The usual interpretation of the type {{b1+b2}} is that we either have
an appropriately tagged value of type {{b1}} or an appropriately
tagged value of type {{b2}}. This interpretation is mainted in the
precense of {{eta+}} and {{eps+}} in the following sense: a value of
type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
switches into a value {{L (-v): -b + b}} that flows backwards on the
upper wire. The inversion of direction is captured by the negative
sign on the value.

\paragraph*{Negative Information Flow.} 
Hence, the intuitive way to read the digrams is that they reverse the
flow of information -- in normal {{langRev}} circuits information
flows from left-to-right and {{eps+}} operations cause information to
flow from right-to-left.  The interpretation of {{eta+}} is precisely
the opposite: it turns a right-to-left flow of information in a
cicruit into a left-to-right flow.

\paragraph*{Time Travelling.}
A time travelling intuition is also applicable here: if view the
left-to-right direction of information flow as the canonical forward
direction of information, then the action of {{eps+}} allows values,
aka information particles, to flow backwards in time. Particles that
flow backward in time get to see and interact with the history of
particles that they coexist with i.e. values that they exist
multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
(whose construction is shown in Section \ref{sec:neg-constructions}).

\paragraph*{Coherence.}
The appropriate coherence condition for Compact Closed Categories,
expressed as the traingle axiom, is satisfied by {{eta+}}/{{eps+}} and
may be visualized as explained by Selinger
\cite{springerlink:10.1007/978-3-642-12821-94} as follows:

@@

\paragraph*{Non-termination.}
As is well known in category theory, Compact Closed Categories contain
trace operations.  In the context of {{langRev}} this implies that
{{eta+}}/{{eps+}} allow the construction of a {{trace+}} operator
(detailed in Section \ref{sec:monoidal-constructions}) hence
introducing \emph{non-termination} into the language
\cite{infeffects}.

\paragraph*{Zero Information.}
The fact that {{eta+}} (and hence {{eps+}}) introduces no information
can be understood is by viewing {{-b+b}} as a type whose value is
observable only if you `pay' {{b}} amount of information to the {{-b}}
branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
amount of information is returned to you. (TODO. more needs to be said
here.)

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics}

\subsection{Syntax}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions in {{langRevEE}} }
\label{sec:monoidal-constructions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions Specific to Negation}
\label{sec:neg-constructions}
